MultifidelityTrajectoryOptimization

### 2025-02-25 artefacts/paper.tex
uce unnecessary sections

\subsection{Mass-Spring-Damper System}
The mass-spring-damper system is a second order system that can be modelled using Newton's second law of motion.
The system in consideration has a mass $m=5$ g, spring constant $k=2$ N/m and damping coefficient $c=0.5$ Ns/m.
Given an external force $F$ and the acceleration of the system $\ddot{x}$,
the dynamics of the system can be written as the equality of forces acting on the mass:

$$
\begin{aligned}
  m\ddot{x} &= F - c\dot{x} - kx \\
  \implies \ddot{x} &= \frac{F}{m} - \frac{c}{m}\dot{x} - \frac{k}{m}x
\end{aligned}
$$

We have the state $[x \dot{x}]^T$ that contains the position and velocity of the mass, both of which are observed.
The control input $u=F$ is applied to the mass, yielding the following state-space model:

@@

Thus, we have a standard continuous-time LTI system $\dot{x} = A_cx + B_cu$ and $y = C_cx$, where

\begin{equation}
  \begin{aligned}
    A_c &=
    \begin{bmatrix}
      0 & 1 \\
      -\frac{k}{m} & -\frac{c}{m}
    \end{bmatrix} \\
    B_c &=
    \begin{bmatrix}
      0 \\
      \frac{1}{m}
    \end{bmatrix} \\
    C_c &=
    \begin{bmatrix}
      1 & 0 \\
      0 & 1
    \end{bmatrix}
  \end{aligned}
\end{equation}

@@

With the resulting state-space model

\begin{equation}
  \begin{aligned}
    x_{k+1}^{ext} &=
    \begin{bmatrix}
      A & B & 0 \\
      0 & \mathbb{I}_{n_u \times n_u} & 0 \\
      0 & 0 & \mathbb{I}_{n_r \times n_r}
    \end{bmatrix}
    \begin{bmatrix}
      x_k \\
      u_{k-1} \\
      r_k
    \end{bmatrix} +
    \begin{bmatrix}
      B \\
      \mathbb{I}_{n_u \times n_u} \\
      0
    \end{bmatrix} \Delta u_k \\
  \end{aligned}
\end{equation}

with

$$
\begin{aligned}
  A^{ext} &=
  \begin{bmatrix}
    A & B & 0 \\
    0 & \mathbb{I}_{n_u \times n_u} & 0 \\
    0 & 0 & \mathbb{I}_{n_r \times n_r}
  \end{bmatrix} \\
  B^{ext} &=
  \begin{bmatrix}
    B \\
    \mathbb{I}_{n_u \times n_u} \\
    0
  \end{bmatrix}
\end{aligned}
$$

@@

$$
\begin{aligned}
  e_k^T Q_e e_k &= (y_k - r_k)^T Q_e (y_k - r_k) \\
  &= \left(
    \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
    \end{bmatrix}
    \begin{bmatrix}
      x_k \\
      u_{k-1} \\
      r_k
  \end{bmatrix}\right)^T Q_e
  \left(
    \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
    \end{bmatrix}
    \begin{bmatrix}
      x_k \\
      u_{k-1} \\
      r_k
  \end{bmatrix}\right) \\
  &= x_k^{ext^T}
  \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
  \end{bmatrix}^T Q_e
  \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
  \end{bmatrix} x_k^{ext} \\
\end{aligned}
$$

\begin{equation}
  Q =
  \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
  \end{bmatrix}^T Q_e
  \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
  \end{bmatrix}
\end{equation}

@@

Using the state transition formula

$$
\begin{aligned}
  x_1 &= A x_0 + B u_0 \\
  x_2 &= A x_1 + B u_1 = A(A x_0 + B u_0) + B u_1 = A^2 x_0 + A B u_0 + B u_1 \\
  \vdots \\
  x_N &= A x_{N-1} + B u_{N-1} = A^{N-1} x_0 + A^{N-2} B u_0 + A^{N-3} B u_1 + \ldots + B u_{N-1} \\
  \implies x_k &= A^k x_0 + \sum_{i=0}^{k-1} A^{k-1-i} B u_i
\end{aligned}
$$

We construct the matrices $X$, $U$, $S$ and $M$ such that $X = SU + Mx_0$.

$$
\begin{aligned}
  X &=
  \begin{bmatrix}
    x_1^{ext} \\
    x_2^{ext} \\
    \vdots \\
    x_N^{ext}
  \end{bmatrix} & \in \mathbb{R}^{Nn_x \times 1} \\
  U &=
  \begin{bmatrix}
    \Delta u_0 \\
    \Delta u_1 \\
    \vdots \\
    \Delta u_{N-1}
  \end{bmatrix} & \in \mathbb{R}^{Nn_u \times 1} \\
  S &=
  \begin{bmatrix}
    B & 0 & \ldots & 0 \\
    AB & B & \ldots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    A^{N-1} B & A^{N-2} B & \ldots & B
  \end{bmatrix} & \in \mathbb{R}^{Nn_x \times Nn_u} \\
  M &=
  \begin{bmatrix}
    A \\
    A^2 \\
    \vdots \\
    A^N
  \end{bmatrix} & \in \mathbb{R}^{Nn_x \times n_x}
\end{aligned}
$$

Similarly, the matrices $\bar{Q}$ and $\bar{R}$ are constructed

$$
\begin{aligned}
  \bar{Q} &=
  \begin{bmatrix}
    Q & 0 & \ldots & 0 \\
    0 & Q & \ldots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & \ldots & 0
  \end{bmatrix} & \in \mathbb{R}^{Nn_x \times Nn_x} \\
  \bar{R} &=
  \begin{bmatrix}
    R & 0 & \ldots & 0 \\
    0 & R & \ldots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & \ldots & R
  \end{bmatrix} & \in \mathbb{R}^{Nn_u \times Nn_u}
\end{aligned}
$$

The objective function can then be written as a quadratic function of the control inputs $U$

@@

\begin{equation}
  \begin{aligned}
    \mathbb{E}[J] &= \mathbb{E}[\sum_{k=0}^{N-1} x_k^T Q x_k + \Delta u_k^T R \Delta u_k] \\
    &= \mathbb{E}[X^T \bar{Q} X + U^T \bar{R} U + x_0^T Q x_0] \\
    &= \mathbb{E}[X^T \bar{Q} X] + \mathbb{E}[U^T \bar{R} U] + \mathbb{E}[x_0^T Q x_0] \\
    &= \mathbb{E}[(S U + M x_0)^T \bar{Q} (S U + M x_0)] + U^T \bar{R} U + \mathbb{E}[x_0^T Q x_0] \\
    &= \mathbb{E}[U^T S^T \bar{Q} S U + U^T S^T \bar{Q} M x_0 + x_0^T M^T \bar{Q} S U + x_0^T M^T \bar{Q} M x_0] + U^T \bar{R} U + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
    &= U^T S^T \bar{Q} S U + \mathbb{E}[U^T S^T \bar{Q} M x_0] + \mathbb{E}[x_0^T M^T \bar{Q} S U] + \mathbb{E}[x_0^T M^T \bar{Q} M x_0] + U^T \bar{R} U + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
    &= U^T (S^T \bar{Q} S + \bar{R}) U + U^T S^T \bar{Q} M \mathbb{E}[x_0] + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T M^T \bar{Q} M x_0] + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
    &= U^T (S^T \bar{Q} S + \bar{R}) U + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] M^T \bar{Q} M \mathbb{E}[x_0] + \text{tr}(M^T \bar{Q} M \text{Cov}[x_0]) \dots \\
    & \quad + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
    &= U^T (S^T \bar{Q} S + \bar{R}) U + 2 \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] (M^T \bar{Q} M + Q) \mathbb{E}[x_0] + \text{tr}((M^T \bar{Q} M + Q) \text{Cov}[x_0]) \\
  \end{aligned}
\end{equation}

@@

In order to formulate the CV and ACV approaches, we need a low-fidelity(LF) model of the system that correlates well
with the high-fidelity(HF) model. The LF model should be computationally cheaper to evaluate than the
HF model, and hence can be used as a control variate to reduce the variance of the objective function
estimation.

@@

\begin{algorithm}
  \caption{Numerical optimization with ACV estimator}
  \begin{algorithmic}
    \Require $J_h, J_l, n_{acv}, m_{acv}, U_0, x_0, k_{max}$
    \State $k \gets 0$ (iteration counter)
    \State $U \gets U_0$ initialize the numerical optimizer
    \While{numerical optimizer has not converged $\wedge$ $k < k_{max}$}
    \State $U_h \gets$ $U$ from the numerical optimizer
    \State $U_{hla}^{(k)} \gets T_{hla} U_h$
    \State $U_{hla}^{max} \gets \underset{U_{hla}(j)} \max \{\rho(J_h(U_h, x_0), J_l(U_{hla}^{(j)}, x_0))\}_{j=1}^k$
    \State $\text{Cov}_{hl} \gets \text{Cov}(J_h(U_h, x_0), J_l(U_{hla}^{max}, x_0))$ calculated statistically using $n_{acv}$ samples
    \State $\text{Var}_l \gets \text{Var}[J_l(U_{hla}^{max}, x_0)]$ calculated statistically using $n_{acv}$ samples
    \State $\alpha \gets -\frac{m_{acv}}{m_{acv} + n_{acv}} \frac{\text{Cov}_{hl}}{\text{Var}_l}$ calculated statistically using $n_{acv}$ samples
    \State $\mathbb{E}[J_l(U_{hla}^{max}, x_0)] \gets \frac{1}{m_{acv}} \sum_{i=n_{acv}}^{n_{acv}+m_{acv}} J_l(U_{hla}^{max}, x_0^{(i)})$
    \State $S_{n_{acv},m_{acv}}^{ACV}(J_h, J_l) \gets S_{n_{acv}}^{MC}(J_h(U_h, x_0)) + \alpha (S_{n_{acv}}^{MC}(J_l(U_{hla}^{max}, x_0)) - S_{m_{acv}}^{MC}(J_l(U_{hla}^{max}, x_0)))$
    \State $k \gets k + 1$
    \State $S_{n_{acv},m_{acv}}^{ACV}(J_h, J_l)$ is passed to the numerical optimizer for computing gradients for the next iteration
    \EndWhile
  \end{algorithmic}
\end{algorithm}

>>>>

\subsection{Mass-Spring-Damper System}
The mass-spring-damper system is a second order system that can be modelled using Newton's second law of motion.
The system in consideration has a mass $m=5$ g, spring constant $k=2$ N/m and damping coefficient $c=0.5$ Ns/m.
Given an external force $F$ and the acceleration of the system $\ddot{x}$, we have the state $[x \dot{x}]^T$
that contains the position and velocity of the mass, both of which are observed.
The control input $u=F$ is applied to the mass, yielding the following state-space model:

@@

Thus, we have a standard continuous-time LTI system $\dot{x} = A_cx + B_cu$ and $y = C_cx$, which can be discretized
into a discrete-time LTI system $x_{k+1} = Ax_k + Bu_k$ and $y_k = Cx_k$ using a sampling time $T_s$. We can now use
this discrete-time system to set up the LQR controller.

@@

And stack the state vectors $x_k^{ext}$ for all time steps $k$ to get the extended state vector $X$.
The control input is also stacked to get the control input vector $U$.

$$
X =
\begin{bmatrix}
  x_1^{ext} \\
  x_2^{ext} \\
  \vdots \\
  x_N^{ext}
\end{bmatrix}  \in \mathbb{R}^{Nn_x \times 1}, \quad
U =
\begin{bmatrix}
  \Delta u_0 \\
  \Delta u_1 \\
  \vdots \\
  \Delta u_{N-1}
\end{bmatrix}  \in \mathbb{R}^{Nn_u \times 1}
$$

@@

@@

Using the state transition formula, we can write the stacked state vector $X$ as $X = SU + Mx_0$.
Here $S \in \mathbb{R}^{Nn_x \times Nn_u}$ and $M \in \mathbb{R}^{Nn_x \times n_x}$ are matrices that depend on the
state transition matrices $A$ and $B$.
Similarly, the matrices $\bar{Q} \in \mathbb{R}^{Nn_x \times Nn_x}$ and $\bar{R} \in \mathbb{R}^{Nn_u \times Nn_u}$
are constructed using the penalty matrices $Q$ and $R$. The objective function then can then be written as a quadratic function of the control inputs $U$

@@

\begin{equation}
  \begin{aligned}
    \mathbb{E}[J] &= \mathbb{E}[\sum_{k=0}^{N-1} x_k^T Q x_k + \Delta u_k^T R \Delta u_k] \\
    &= U^T (S^T \bar{Q} S + \bar{R}) U + 2 \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] (M^T \bar{Q} M + Q) \mathbb{E}[x_0] + \text{tr}((M^T \bar{Q} M + Q) \text{Cov}[x_0]) \\
  \end{aligned}
\end{equation}

@@

The solution to the stochastic LQR controller, $U^*$ can be written as a function of the initial state $x_0$.

$$
\begin{aligned}
  U^* &= -H^{-1} q \\
  \mathbb{E}[U^*] &= \mathbb{E}[-H^{-1} q] \\
  &= -H^{-1} \mathbb{E}[q] \text{ ($H$ is deterministic)} \\
  &= -H^{-1} \mathbb{E}[S^T \bar{Q} M x_0] \\
  &= -H^{-1} S^T \bar{Q} M \mathbb{E}[x_0] \\
  \text{Cov}[U^*] &= \text{Cov}[-H^{-1} q] \\
  &= H^{-1} \text{Cov}[q] H^{-1} \text{ ($H$ is symmetric)} \\
  &= H^{-1} \text{Cov}[S^T \bar{Q} M x_0] H^{-1} \\
  &= H^{-1} S^T \bar{Q} M \text{Cov}[x_0] M^T \bar{Q} S H^{-1} \text{($\bar{Q}$ is symmetric)} \\
\end{aligned}
$$

Thus, we aim to reduce the variance of the objective function estimate thereby reducing the covariance of the
solution $U^*$ given a finite cost of computation.

In order to formulate the CV and ACV approaches, we need a low-fidelity(LF) model of the system that correlates well
with the high-fidelity(HF) model. The LF model should be computationally cheaper to evaluate than the
HF model, and hence can be evaluated with a lot more samples of $x_0$ than the HF model for the same cost.
Thus, we use the LF model as a control variate to reduce the variance of the objective function
estimation.

@@

\begin{algorithm}
  \caption{Numerical optimization with ACV estimator}
  \begin{algorithmic}
    \Require $J_h, J_l, n_{acv}, m_{acv}, U_0, x_0, k_{max}$
    \State $k \gets 0$ (iteration counter)
    \State $U \gets U_0$ initialize the numerical optimizer
    \While{numerical optimizer has not converged $\wedge$ $k < k_{max}$}
    \State $U_h \gets$ $U$ from the numerical optimizer
    \State $U_{hla}^{(k)} \gets T_{hla} U_h$
    \State $U_{hla}^{max} \gets \underset{U_{hla}(j)} \max \{\rho(J_h(U_h, x_0), J_l(U_{hla}^{(j)}, x_0))\}_{j=1}^k$
    \State $\text{Cov}_{hl} \gets \text{Cov}(J_h(U_h, x_0), J_l(U_{hla}^{max}, x_0))$ calculated statistically using $n_{acv}$ samples
    \State $\text{Var}_l \gets \text{Var}[J_l(U_{hla}^{max}, x_0)]$ calculated statistically using $n_{acv}$ samples
    \State $\alpha \gets -\frac{m_{acv}}{m_{acv} + n_{acv}} \frac{\text{Cov}_{hl}}{\text{Var}_l}$ calculated statistically using $n_{acv}$ samples
    \State $S_{m_{acv}}^{MC}(J_l(U_{hla}^{max}, x_0)) \gets \frac{1}{m_{acv}} \sum_{i=n_{acv}}^{n_{acv}+m_{acv}} J_l(U_{hla}^{max}, x_0^{(i)})$
    \State $S_{n_{acv},m_{acv}}^{ACV}(J_h, J_l) \gets S_{n_{acv}}^{MC}(J_h(U_h, x_0)) + \alpha (S_{n_{acv}}^{MC}(J_l(U_{hla}^{max}, x_0)) - S_{m_{acv}}^{MC}(J_l(U_{hla}^{max}, x_0)))$
    \State $k \gets k + 1$
    \State $S_{n_{acv},m_{acv}}^{ACV}(J_h, J_l)$ is passed to the numerical optimizer for computing gradients for the next iteration
    \EndWhile
  \end{algorithmic}
\end{algorithm}
